It's easier to move sideways

### 2023-09-07 manuscript.Rmd
again, a ton of updaes

library(tidyverse)
library(pander)
library(fs)
library(here)
library(english)
library(phyloseq) 
library(vegan)
library(ggrepel)
library(here)
library(worrms)
```

@@

format_species <- function(spp) {
  if (spp$found) {
    str_glue("*{spp$species}* {spp$authority}")
  } else {
    s <- str_split(spp$species," ")[[1]]
    str_glue("*{s[1]}* {s[2]}")
  }
}

@@

permanova <- function(modl,d=1) {
  m <- suppressWarnings(broom::tidy(modl)) %>%
    filter(!term %in% c("Total","Residual"))
  if (nrow(m) > 1) {
    m %>%
      pmap_chr(~{
        with(list(...), {
          str_glue("{term}: DF = {round(df,1)}, SS = {round(SumOfSqs,1)}, $\\mathrm{{R}}^2$ = {round(R2,1)}, pseudo-F = {round(statistic,1)}, {pval(p.value)}")
        })
      }) %>%
      str_c(collapse = "; ")
  } else {
    with(m %>% slice(1),{
      str_glue("DF = {round(df,1)}, SS = {round(SumOfSqs,1)}, $\\mathrm{{R}}^2$ = {round(R2,1)}, pseudo-F = {round(statistic,1)}, {pval(p.value)}")
    }) 
  }
}

@@

print_table <- function(tbl,id="",caption="",hdr_case="sentence") {
  if (!grepl("^: ",caption)) {
    caption <- paste0(": ",caption)
  }

@@

@@

dir_create(here("cache"))
if (file_exists(here("cache","all_summary.rds"))) {
  all_summary <<- read_rds(here("cache","all_summary.rds"))
} else {
  all_summary <<- sample_summary(cc$crabs_untransformed)
  write_rds(all_summary,here("cache","all_summary.rds"))
}
if (file_exists(here("cache","deep_summary.rds"))) {
  deep_summary <<- read_rds(here("cache","deep_summary.rds"))
} else {
  deep_summary <<- sample_summary(cc$crabs_deep)
  write_rds(deep_summary,here("cache","deep_summary.rds"))
}

@@

@@

A total of 89 ARMS units were deployed over the course of six years (2010--2016). Units were deployed by NOAA scuba divers during Pacific Reef Assessment and Monitoring Program (RAMP) research cruises in shallow coral reef habitat (12–16 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@tbl:arms_table]). The sites in the MHI were at Hawai&#x02BB;i Island, Maui, O&#x02BB;ahu, and Kaua&#x02BB;i. The sites in the NWHI were at Lalo (French Frigate Shoals), Kapou (Lisianski), Manawai (Pearl & Hermes), and Hōlanikū (Kure Atoll) ([@fig:site_map]). To explore the effects of depth on crab community structure, we also present data from a previous study by Hurley et al. [-@hurleyAssessmentShallowMesophotic2016], which included a total of 27 units deployed off O&#x02BB;ahu from 2009–2012 in 12–90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]). Shallow units were deployed according to standard NOAA methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed-gas closed-circuit rebreather diving.

@@

```{r site-map, fig.cap='Sampling map (Papahānaumokuākea Marine National Monument shaded in light blue)', results="asis", dev="svg", fig.width=9, fig.height=6, dependson="init"}
library(sf)
library(rnaturalearth)

@@

ggplot() +
  # geom_contour(data=bath_proj,mapping=aes(x=V1,y=V2,z = V3),  color="#cecece") +
  # geom_raster(data=bpr,aes(x=x,y=y,fill=z)) +
  # geom_tile(data=bath_proj,mapping=aes(x=V1,y=V2,fill = V3)) +
  # scale_fill_gradient2(low="dodgerblue4", mid="gainsboro", high="darkgreen") +
  geom_sf(data=hawaii %>% st_transform(transformer), color="black", fill="green", linewidth=0.4) +
  geom_sf(data=pmnm,fill="#6ed8f6",alpha=0.1) +
  geom_sf(data=crab_sites %>% st_transform(transformer),aes(fill=hawaiian_name),size=3,shape=21,color="black") +
  scale_fill_manual(values=island_pal,name="Island") +
  coord_sf(xlim=box_after[c(1,3)],ylim=box_after[c(2,4)]) +
  scale_x_continuous(breaks = -seq(180, 155, -5)) +
  theme_minimal() +
  xlab("Longitude") +
  ylab("Latitude") +
  theme(legend.text = element_text(size=10),legend.title = element_text(size=12))
cat("{#fig:site_map}")
```

@@

We used the distance-based redundancy analysis method (dbRDA; a form of multivariate multiple regression) implemented in the `dbrda` function from vegan to examine the relative importance of environmental variables on crab community variation across the Hawaiian Islands. We first assessed the importance of each environmental variable individually using marginal tests. Then, a stepwise forward selection procedure (using the function `ordiR2step`) was used to find the suite of environmental variables that best explained the dissimilarity among sites with regard to crab community composition. Colinearity between traits was assessed by calculating variance inflation factors (VIF) using the function `vif.cca` from complete and reduced models and discarding variables where VIF > 10. Significance levels were assessed with the function `anova` using permutation tests with 9,999 permutations. All continuous environmental variables used for dbRDA were scaled to unit variance. We used the permutational analysis of variance (PERMANOVA) method in the `adonis2` function from vegan to assess variation among sites and/or depths across the islands, with 9,999 permutations to assess significance in both term-based and marginal tests.

@@

```{r env-table,results="asis", dependson="init"}
print_table(here("data","tables","environmental_variables.csv"),"env_table","Potential environmental drivers of crab community structure, assembled from PacIOOS and @selkoeDNACoralReef2016")
```

@@

The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs. Sampling units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals per unit, and abundance ranged from `r format_range(all_summary$abundance_range)` individuals. Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct types (including `r all_summary$types_breakdown$species` species and `r all_summary$types_breakdown$morphospecies` morphospecies) in `r all_summary$taxonomy_breakdown$genera` genera and `r all_summary$taxonomy_breakdown$families` families, which accounted for ~36% of the known crab diversity reported from Hawaiian waters [@castroCatalogAnomuranBrachyuran2011]. The four most common species were: `r with(all_summary,top_list(top5,top5_n,top5_units,fig="fig_crabs"))`. `r str_to_sentence(as.english(all_summary$singleton_count))` species occurred in a single unit across all sites (singletons), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.

![The four most abundant crab species found across all deployed ARMS units. (A) *Chlorodiella laevissima*. (B) *Perinia tumida*. (C) *Percnon abbreviatum*. (D) *Dynomene hispida*.](data/images/4crabs.png){#fig:fig_crabs}

@@

The `r nsamples(cc$crabs_deep)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded both fewer species and fewer individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep)),1)` crabs overall, a total of `r sum(unlist(deep_summary$types_breakdown))` distinct types (`r deep_summary$types_breakdown$species` species and `r deep_summary$types_breakdown$morphospecies` morphospecies), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)` crabs. The four most abundant species in the deep ARMS did not overlap with those from the complete dataset: `r with(deep_summary,top_list(top5,top5_n,top5_units))`. There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r num(deep_summary$singleton_count/ntaxa(cc$crabs_deep)*100,1)`% of species found on deep ARMS.

@@

For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different, although mean values were similar with `r num(all_alpha$richness_main$richness,0.1)` $\pm$ `r num(all_alpha$richness_main$richness_sd,0.1)` in the MHI and `r num(all_alpha$richness_nwhi$richness,0.1)` $\pm$ `r num(all_alpha$richness_nwhi$richness_sd,0.1)` in the NWHI ([@fig:alpha_box]; `r t_test(all_alpha$t_s_island,plain=TRUE)`). When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`).

```{r alpha-box, fig.cap='Box plots of Simpson diversity by island group. Notches show nonparametric confidence intervals about the median', results="asis", dev="svg", fig.width=10, fig.height=8, dependson="init"}
ggplot(all_alpha$richness_table %>% mutate(island_group = str_to_sentence(island_group))) + 
  geom_boxplot(aes(x=island_group,y=Simpson,fill=island_group),notch=TRUE) +
  scale_fill_manual(values=island_group_pal) +
  guides(fill="none") + 
  theme_bw() + 
  theme(panel.grid = element_blank()) + 
  xlab("Island group") + 
  ylab("Simpson's diversity index") 
cat("{#fig:alpha_box}")
```

@@

```{r permanova, dependson="init"}
# PERMANOVA of 
anova_region <- adonis2(cc$crab_dist_shallow ~ island_group,data=cc$crab_data_shallow,permutations=9999)
anova_island <- adonis2(cc$crab_dist_shallow ~ island,data=cc$crab_data_shallow,permutations=9999)
```

There were significant differences in shallow crab community structure between the Main (MHI) and Northwest Hawaiian Islands (`adonis2` PERMANOVA: `r permanova(anova_region)`) as well as among individual islands (`r permanova(anova_island)`; see [@fig:pairwise_island] for results of pairwise comparisons).

@@

```{r pairwise-island, fig.cap='Results of pairwise PERMANOVA tests by island (FDR-adjusted p-values)', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
library(ggtext)
pairwise_island <- pairwise_adonis(cc$crab_dist_shallow,cc$crab_data_shallow$island,permutations=999)

@@

pw1 <- pairwise_island %>%
  select(left,right,p_adj) %>%
  mutate(left=factor(left,levels=rev(islands)),right=factor(right,levels=rev(islands))) %>%
  mutate(sig = p_adj < 0.05, p_adj = pval(p_adj)) %>%
  mutate(
    combo = map2_chr(left,right,~str_c(sort(c(.x,.y)),collapse="-"))
  ) %>%
  select(-c(left,right)) %>%
  separate_wider_delim(combo,"-",names=c("left","right")) %>%
  select(left,right,everything()) %>%
  mutate(left=factor(left,levels=islands),right=factor(right,levels=islands))

ggplot(pw1) + 
  # geom_tile(aes(x=right,y=left,fill=sig),width=0.6,height=0.8) + 
  # geom_tile(aes(x=right,y=left),width=0.6,height=0.8) + 
  geom_richtext(
    aes(
      x=right,
      y=left,
      label=p_adj,
      fontface = if_else(sig,"bold","plain")
    ),
    fill = NA,
    label.color = NA
  ) +
  theme_bw() + 
  theme(
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color="black")
  ) +
  # scale_fill_manual(values = c("white","lightgoldenrod")) +
  guides(fill="none") +
  xlab("") + ylab("") +
  scale_x_discrete(position="top") +
  scale_y_discrete(limits=rev)
cat("{#fig:pairwise_island}")
```

>>>>

library(tidyverse)
library(pander)
library(fs)
library(here)
library(english)
library(phyloseq) 
library(vegan)
library(ggrepel)
library(here)
library(worrms)
library(BiodiversityR)
```

@@

plotz_theme <- function() {
  theme_bw() + 
    theme(
      legend.key = element_blank(),
      panel.grid = element_blank()
    )
}

format_species <- function(...) UseMethod("format_species")

format_species.list <- function(spp,authority=TRUE,...) {
  if (spp$found) {
    a <- ifelse(authority,str_glue(" {spp$authority}"),"")
    str_glue("*{spp$species}*{a}")
  } else {
    s <- str_split(spp$species," ")[[1]]
    str_trim(str_glue("*{s[1]}* {s[2]}"))
  }
}

format_species.character <- function(species,found,authority,auth=TRUE,...) {
  spp <- map2_chr(species,found,~{
    .y <- ifelse(is.na(.y),FALSE,.y)
    s <- list(species=.x,found=.y,authority="")
    format_species(s,FALSE)
  })
  if (auth) {
    authority <- if_else(!is.na(authority),str_c(" ",authority),"")
    spp <- str_c(spp,authority)
  }
  return(spp)
}

format_species.data.frame <- function(spp,authority=TRUE,...) {
  spp %>%
    pwalk(~{
      row <- list(...)
      # s <- list(found = row$found, authority=row$authority,otu=row$otu)
      s <- row[c('species','found','authority','otu')]
      print(format_species(s,authority))
    })
}

@@

permdisp <- function(bd,d=1,perm=999) {
  a <- anova(bd,permutations=perm) %>%
    broom::tidy() %>%
    filter(term == "Groups")
  with(a,str_glue("DF = {round(df,d)}, SS = {round(sumsq,d)}, F = {round(statistic,d)}, {pval(p.value)}"))
}

permanova <- function(modl,d=1,stat="pseudo-F",by=NULL) {
  has_r2 <- TRUE
  if (is(modl,"dbrda")) {
    modl <- anova(modl,by=by)
    has_r2 <- FALSE
  }
  m <- suppressWarnings(broom::tidy(modl)) %>%
    filter(!term %in% c("Total","Residual"))
  if (nrow(m) > 1) {
    m %>%
      pmap_chr(~{
        with(list(...), {
          if (has_r2) {
            r2 <- str_glue(" $\\mathrm{{R}}^2$ = {round(R2,d)},") 
          } else {
            r2 <- ""
          }
          str_glue("{term}: DF = {round(df,d)}, SS = {round(SumOfSqs,d)},{r2} {stat} = {round(statistic,d)}, {pval(p.value)}")
        })
      }) %>%
      str_c(collapse = "; ")
  } else {
    with(m %>% slice(1),{
      if (has_r2) {
        r2 <- str_glue(" $\\mathrm{{R}}^2$ = {round(R2,d)},") 
      } else {
        r2 <- ""
      }
      str_glue("{term}: DF = {round(df,d)}, SS = {round(SumOfSqs,d)},{r2} {stat} = {round(statistic,d)}, {pval(p.value)}")
      # str_glue("DF = {round(df,d)}, SS = {round(SumOfSqs,d)}, $\\mathrm{{R}}^2$ = {round(R2,d)}, pseudo-F = {round(statistic,d)}, {pval(p.value)}")
    }) 
  }
}

@@

print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0)) {
  if (!grepl("^: ",caption)) {
    caption <- paste0(": ",caption)
  }

@@

  tbl <- tbl %>%
    select(-any_of(hide))

@@

dir_create(here("cache"))
if (file_exists(here("cache","all_summary.rds"))) {
  all_summary <<- read_rds(here("cache","all_summary.rds"))
} else {
  all_summary <<- sample_summary(cc$crabs_untransformed,4)
  write_rds(all_summary,here("cache","all_summary.rds"))
}
if (file_exists(here("cache","deep_summary.rds"))) {
  deep_summary <<- read_rds(here("cache","deep_summary.rds"))
} else {
  deep_summary <<- sample_summary(cc$crabs_deep,4)
  write_rds(deep_summary,here("cache","deep_summary.rds"))
}

@@

# colors for sample group / depth zone
sample_depth_pal <- c(
  "shallow" = "#ffb400",
  "Shallow (MHI)" = "#ffb400",
  "shallow_main" = "#ffb400",
  "Shallow (NWHI)" = "#9080ff",
  "shallow_nwhi" = "#9080ff",
  "30 m" = "#d87939",
  "30m" = "#d87939",
  "60 m" = "#04c3c8",
  "60m" = "#04c3c8",
  "90 m" = "#04738d",
  "90m" = "#04738d",
  "deep" = "#04738d"
)

PERM <<- 999

@@

A total of 89 ARMS units were deployed over the course of six years (2010--2016). Units were deployed by NOAA scuba divers during Pacific Reef Assessment and Monitoring Program (RAMP) research cruises in shallow coral reef habitat (12–16 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@tbl:arms_table]). The sites in the MHI were at Hawai&#x02BB;i Island, Maui, O&#x02BB;ahu, and Kaua&#x02BB;i. The sites in the NWHI were at Lalo (French Frigate Shoals), Kapou (Lisianski), Manawai (Pearl & Hermes), and Hōlanikū (Kure Atoll) ([@fig:fig_map]). To explore the effects of depth on crab community structure, we also present data from a previous study by Hurley et al. [-@hurleyAssessmentShallowMesophotic2016], which included a total of 27 units deployed off O&#x02BB;ahu from 2009–2012 in 12–90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]). Shallow units were deployed according to standard NOAA methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed-gas closed-circuit rebreather diving.

@@

```{r site-map, fig.cap='Sampling map (Papahānaumokuākea Marine National Monument shaded in light blue)', results="asis", fig.width=9, fig.height=6, dependson="init"}
library(sf)
library(rnaturalearth)

@@

ggplot() +
  # geom_contour(data=bath_proj,mapping=aes(x=V1,y=V2,z = V3),  color="#cecece") +
  # geom_raster(data=bpr,aes(x=x,y=y,fill=z)) +
  # geom_tile(data=bath_proj,mapping=aes(x=V1,y=V2,fill = V3)) +
  # scale_fill_gradient2(low="dodgerblue4", mid="gainsboro", high="darkgreen") +
  geom_sf(data=hawaii %>% st_transform(transformer), color="black", fill="green", linewidth=0.4) +
  geom_sf(data=pmnm,fill="#6ed8f6",alpha=0.1) +
  geom_sf(data=crab_sites %>% st_transform(transformer),aes(fill=hawaiian_name),size=3,shape=21,color="black") +
  scale_fill_manual(values=island_pal,name="Island") +
  coord_sf(xlim=box_after[c(1,3)],ylim=box_after[c(2,4)]) +
  scale_x_continuous(breaks = -seq(180, 155, -5)) +
  theme_minimal() +
  xlab("Longitude") +
  ylab("Latitude") +
  theme(legend.text = element_text(size=10),legend.title = element_text(size=12))
cat("{#fig:fig_map}")
```

@@

We used the distance-based redundancy analysis method (dbRDA; a form of multivariate multiple regression) implemented in the `dbrda` function from vegan to examine the relative importance of environmental variables on crab community variation across the Hawaiian Islands. We first assessed the importance of each environmental variable individually using marginal tests. Then, a stepwise forward selection procedure (using the function `ordiR2step`) was used to find the suite of environmental variables that best explained the dissimilarity among sites with regard to crab community composition. Colinearity between traits was assessed by calculating variance inflation factors (VIF) using the function `vif.cca` from complete and reduced models and discarding variables where VIF > 10. Significance levels were assessed with the function `anova` using permutation tests with `r num(PERM,1)` permutations. All continuous environmental variables used for dbRDA were scaled to unit variance. We assessed community variation among groups using the `betadisper` and `adonis2` functions from vegan. These functions implement analysis of multivariate homogeneity of group dispersions (PERMDISP) and multivariate permutational analysis of variance (PERMANOVA), respectively. The PERMDISP method is used to assess variability within groups and PERMANOVA is used to assess differences among groups. We assessed significance with permutational tests, using `r num(PERM,1)` permutations. Crab community variation was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in the `BiodiversityR` package [@kindtTreeDiversityAnalysis2005].

@@

```{r env-table,results="asis", dependson="init"}
print_table(here("data","tables","environmental_variables.csv"),"env_table","Potential environmental drivers of crab community structure, assembled from PacIOOS and @selkoeDNACoralReef2016",hide="variable")
```

@@

The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs. Sampling units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals per unit, and abundance ranged from `r format_range(all_summary$abundance_range)` individuals. Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct types (including `r all_summary$types_breakdown$species` species and `r all_summary$types_breakdown$morphospecies` morphospecies) in `r all_summary$taxonomy_breakdown$genera` genera and `r all_summary$taxonomy_breakdown$families` families, which accounted for ~36% of the known crab diversity reported from Hawaiian waters [@castroCatalogAnomuranBrachyuran2011]. The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`. `r str_to_sentence(as.english(all_summary$singleton_count))` species occurred in a single unit across all sites (singletons), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.

![The four most abundant crab species found across all deployed ARMS units. (A) *Chlorodiella laevissima*. (B) *Perinia tumida*. (C) *Percnon abbreviatum*. (D) *Dynomene hispida*. Photo credit: Florida Museum of Natural History, licensed [CC BY-NC 3.0](https://creativecommons.org/licenses/by-nc/3.0/)](data/images/4crabs.png){#fig:fig_crabs}

@@

The `r nsamples(cc$crabs_deep)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded both fewer species and fewer individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep)),1)` crabs overall, a total of `r sum(unlist(deep_summary$types_breakdown))` distinct types (`r deep_summary$types_breakdown$species` species and `r deep_summary$types_breakdown$morphospecies` morphospecies), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)` crabs. The four most abundant species in the deep ARMS did not overlap with those from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`. There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r num(deep_summary$singleton_count/ntaxa(cc$crabs_deep)*100,1)`% of species found on deep ARMS.

@@

For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different, although mean values were similar with `r num(all_alpha$richness_main$richness,0.1)` $\pm$ `r num(all_alpha$richness_main$richness_sd,0.1)` in the MHI and `r num(all_alpha$richness_nwhi$richness,0.1)` $\pm$ `r num(all_alpha$richness_nwhi$richness_sd,0.1)` in the NWHI ([@fig:fig_alpha]; `r t_test(all_alpha$t_s_island,plain=TRUE)`). When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`).

```{r alpha-box, fig.cap='Box plots of Simpson diversity by island group. Notches represent nonparametric confidence intervals about the median', results="asis", dev="svg", fig.width=10, fig.height=8, dependson="init"}
ggplot(all_alpha$richness_table %>% mutate(island_group = str_to_sentence(island_group))) + 
  geom_boxplot(aes(x=island_group,y=Simpson,fill=island_group),notch=TRUE) +
  scale_fill_manual(values=island_group_pal) +
  guides(fill="none") + 
  plotz_theme() + 
  xlab("Island group") + 
  ylab("Simpson's diversity index") 
cat("{#fig:fig_alpha}")
```

@@

```{r permanova, dependson="init"}
# PERMANOVA of 
anova_region <- adonis2(cc$crab_dist_shallow ~ island_group,data=cc$crab_data_shallow,permutations=PERM)
anova_island <- adonis2(cc$crab_dist_shallow ~ island,data=cc$crab_data_shallow,permutations=PERM)
bd_region <- betadisper(cc$crab_dist_shallow,cc$crab_data_shallow$island_group)

anova_depth <- adonis2(cc$crab_dist ~ depth_zone,data=cc$crab_data,permutations=PERM)
bd_depth <- betadisper(cc$crab_dist,cc$crab_data$depth_zone)

anova_both <- adonis2(cc$crab_dist ~ island_group + depth_zone, data=cc$crab_data,permutations = PERM, by="margin")
effects <- broom::tidy(anova_both) %>%
  filter(!term %in% c("Residual","Total")) %>%
  pull(R2)
effect_ratio <- max(effects)/min(effects)

cd <- cc$crabs_deep %>%
  sample_tibble()
bd_deep <- betadisper(distance(cc$crabs_deep,"bray"),cd$depth_zone)
```

```{r rda-analysis, results="hide", dependson="init"}
db_upr <- dbrda(cc$crab_otus_shallow ~ depth + chl + sst + slope + coral_cover + closest_island + larval_connectivity + human_impact,data=cc$crab_data_shallow,distance="bray")
db_lwr <- dbrda(cc$crab_otus_shallow  ~ 1,data=cc$crab_data_shallow,distance="bray")

# do our forward selection to find the best model
# set trace to TRUE if you want to see a bunch of output
db_modl <- ordiR2step(db_lwr,db_upr,trace=FALSE)

# see marginal effects of model terms
# anova(db_modl,by="margin")

# vif.cca(db_upr)
vif.cca(db_modl)

# warning: the following *may* be sketchy:
# dbRDA doesn't automatically give you species scores, so let's assign them
# but the community data needs to be transformed to make sense
# see https://github.com/vegandevs/vegan/issues/254 for information on the transformation below
sppscores(db_modl) <- sqrt(decostand(cc$crab_otus_shallow, "total")/2)

# this was to grab the 4 most abundant species overall
top_species <- names(head(rev(sort(taxa_sums(cc$crabs_shallow))),n=4))

# extract the species scores for the species we want
species_scores <- scores(db_modl,display="species") %>%
  as_tibble(rownames = "species") %>%
  filter(species %in% top_species)

model_terms <- read_csv(here("data","tables","environmental_variables.csv")) %>%
  select(variable,driver=`Environmental Driver`)
rda_table <- anova(db_modl,by="term") %>%
  broom::tidy() %>%
  inner_join(model_terms,by=c("term" = "variable")) %>%
  # rename(Term=driver) %>%
  select(-term) %>%
  select(Term=driver,DF=df,`Sum of Squares`=SumOfSqs,`F-statistic`=statistic,`p-value`=p.value)

abs_r2 <- diff(c(0,db_modl$anova$R2.adj))
```
### Environmental drivers {-}

The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r scales::percent(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_rda]). Seven of the 10 environmental variables tested were significant ([@tbl:rda_table]). The drivers contributing most to the explained variation were chlorophyll-A ($\mathrm{R}^2$ = `r num(abs_r2[1],0.01)`) and human impacts ($\mathrm{R}^2$ = `r num(abs_r2[2],0.01)`).

```{r rda, fig.cap="Distance-based redundancy analysis (dbRDA) plot showing significant environmental drivers of crab community variation across the Hawaiian Islands.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
rda_plotz <- plot_ord(cc$crabs_shallow,db_modl,scale=2,color="island_group") + 
  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color=NA,alpha=0.1) +
  geom_point(aes(fill=island_group),shape=21,color="black",size=3) +
  plotz_theme() + 
  scale_fill_manual(values=island_group_pal)

xvar <- eigenvals(db_modl)/sum(eigenvals(db_modl))
r2 <- RsquareAdj(db_modl)$adj.r.squared
constrained <- xvar*r2

splained <- function(name,constrained,total,accuracy=0.1) {
  str_glue("{name} ({scales::percent(constrained,accuracy)} constrained, {scales::percent(total,accuracy)} total)")
}

rda_plotz + 
  xlab(splained("dbRDA1",xvar[1],constrained[1])) +
  ylab(splained("dbRDA2",xvar[2],constrained[2])) 
cat("{#fig:fig_rda}")
```

### Community variation {-}

```{r cap-analysis,dependson="init",echo=FALSE,message=FALSE,warning=FALSE,results="hide"}
# convert otu tables and sample data into appropriate format
otus <- otu_table(cc$crabs) %>%
  as.data.frame()
cd <- cc$crab_data %>%
  mutate(sample_group = factor(sample_group,ordered=FALSE)) %>%
  as.data.frame()

# run CAP analysis, taking care of negative eigenvalues where necessary
cap_depth <- CAPdiscrim(otus ~ sample_group, data=cd, dist="bray", axes=2, m=0, add=TRUE)
# # calculate proportion of explained variance
cap_var <- cap_depth$manova$Eigenvalues / sum(cap_depth$manova$Eigenvalues)
cap_var <- list(x = cap_var[1], y = cap_var[2])
```

Shallow crab community structure was significantly different between the MHI and NWHI (`r permanova(anova_region,2)`) as well as among individual islands (`r permanova(anova_island,2)`; see [@fig:fig_pairwise] for results of pairwise comparisons). There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`). When mesophotic ($\geq$ 30 m) units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) (`r permanova(anova_depth,2)`) and community dispersion (`r permdisp(bd_depth)`). However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`). Although crab communities varied significantly among both depth zones and island groups (`r permanova(anova_both,2)`), the effect size (model $\mathrm{R}^2$) of depth was `r num(effect_ratio,0.1)` times greater than that of island group. In the CAP analysis, samples were successfully assigned to their respective sampling groups `r scales::percent(cap_depth$percent/100,0.1)` of the time. The first two axes explained `r scales::percent(cap_var$x,0.1)` and `r scales::percent(cap_var$y,0.1)` of the variation, respectively. The samples from the three mesophotic groups on O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km of the Hawaiian Islands ([@fig:fig_cap]). In addition to grouping apart from shallow samples, there was as much or more variation across the 30--90 m samples than across all of the shallow samples, although that variation was largely due to the differences between the 30 m samples and the 60 m and 90 m samples ([@fig:fig_cap]).

```{r cap, fig.cap="Canonical analysis of principal coordinates (CAP) plot showing variation in brachyuran crab communities by sampling region and depth zone. Ellipses represent 95% confidence intervals for sample groups.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
# make the sample group look nice
nice <- function(s) {
  str_replace(s,"shallow","Shallow") %>%
    str_replace("main","(MHI)") %>%
    str_replace("nwhi","(NWHI)") %>%
    str_split("((?<=[0-9])(?=[a-zA-Z]))|(_)") %>%
    map_chr(~str_c(.x,collapse=" "))
}

# get all the stuff we need to plot
pcoa <- cap_depth$x %>%
  as_tibble(rownames="sample") %>%
  left_join(cd,by="sample") %>%
  mutate(
    sample_group = nice(sample_group),
    sample_group = fct_reorder(sample_group,depth)
  )

ggplot(pcoa,aes(x=LD1,y=LD2,fill=sample_group)) + 
  stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
  geom_point(color="black",shape=21,size=5)  +
  # scale_color_manual(values=sample_depth_pal) + 
  scale_fill_manual(values=sample_depth_pal,name="Sample group") +
  xlab(str_glue("Linear discriminant 1 [{scales::percent(cap_var$x,0.1)}]")) +
  ylab(str_glue("Linear discriminant 2 [{scales::percent(cap_var$y,0.1)}]")) +
  plotz_theme()
cat("{#fig:fig_cap}")
```

<!-- 
Here's the thing: it's not that the differences between MHI & NWHI become non-significant with the inclusion of deep samples, 
it's just that the deep samples are MORE different from all the shallow samples than all the shallow samples are from one another?
We can think of PERMANOVA R^2 and ANOSIM R as effect sizes. Thus, it may be a bit weak but for shallow samples only, island group 
R2 = 0.072, for all samples R2 = 0.065. For depth (all samples of course), R2 = 0.110. If we consider the effect sizes from ANOSIM, 
for shallow samples, island group R = 0.175, for all samples, island group R = 0.161, and for shallow vs deep,  R = 0.509. 
The apparent difference in effect size is much more pronounced in the ANOSIM, but it's different in each case, and it's smaller
for all samples than it is for just shallow samples.
-->

@@

```{r pairwise-island, fig.cap='Significance levels of pairwise PERMANOVA tests by island (FDR-adjusted p-values)', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
library(ggtext)
pairwise_island <- pairwise_adonis(cc$crab_dist_shallow,cc$crab_data_shallow$island,permutations=PERM)

@@

pw1 <- pairwise_island %>%
  select(left,right,p_adj) %>%
  mutate(left=factor(left,levels=rev(islands)),right=factor(right,levels=rev(islands))) %>%
  mutate(sig = round(p_adj,2) < 0.05, p_adj = pval(p_adj)) %>%
  mutate(
    combo = map2_chr(left,right,~str_c(sort(c(.x,.y)),collapse="-"))
  ) %>%
  select(-c(left,right)) %>%
  separate_wider_delim(combo,"-",names=c("left","right")) %>%
  select(left,right,everything()) %>%
  mutate(left=factor(left,levels=islands),right=factor(right,levels=islands))

ggplot(pw1) + 
  geom_richtext(
    aes(
      x=right,
      y=left,
      label=p_adj,
      fontface = if_else(sig,"bold","plain")
    ),
    fill = NA,
    label.color = NA
    # size=10
  ) +
  plotz_theme() +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(color="black"),
    axis.text = element_text(size = 14),
    axis.text.x = element_text(angle=30,vjust=1,hjust=0)
  ) +
  guides(fill="none") +
  xlab("") + ylab("") +
  scale_x_discrete(position="top") +
  scale_y_discrete(limits=rev)
cat("{#fig:fig_pairwise}")
```

```{r rda-table,results="asis",dependson="rda-analysis"}
print_table(rda_table,"rda_table","Significant terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none") 
```
